---
title: "Bayesian Linear Regression"
subtitle: "Analysis of Flight Delay Data"
author: "Sara Parrish, Heather Anderson (Advisor: Dr. Seals)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
date-format: "MMM D, YYYY"
format:
  revealjs:
    css: custom.css
    theme: dark
    embed-resources: true
    slide-number: false
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
course: Capstone Projects in Data Science
bibliography: references.bib
csl: ieee.csl
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
pdf-separate-fragments: true
fig-align: center
---


::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::


## Introduction to Bayesian Linear Regression

- Regression under the *frequentist* framework
    - Independent variables are used to predict dependent variables
    - Linear regression finds best-fitting line to observed data to make further predictions
      - Regression parameters ($\beta$) are assumed to be fixed
    - Only collected data is used for approximation
- Regression under the *Bayesian* framework
    - Independent variables are used to predict dependent variables
    - Regression parameters ($\beta$) *are not* assumed to be fixed
    - Collected data is used alongside prior knowledge for approximation

## Introduction  {.smaller}

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

# Methods

## Frequentist vs. Bayesian Approach

- The **Frequentist** Approach
  - Typical linear model 
  
$$
Y = \beta_0 + \beta_1X + \varepsilon
$$

  - $Y$ : Dependent variable, the outcome
  - $\beta_0$ : y intercept
  - $\beta_1$ : The regression coefficient 
  - $X$ : Independent variable
  - $\varepsilon$ : Random error [@xiaogang]
- $\hat\beta$ provides a point estimate


## Frequentist vs. Bayesian Approach

- The **Bayesian** Approach
  - A regression is constructed using probability distributions, not point estimates as in the frequentist approach
  - Bayes Rule (2) is used to inform the model [@xiaogang]

$$
p(B|A) = \frac{p(A|B)\cdot p(B)}{p(A)}
$$

- Bayes Rule allows for the calculation of inverse probability ($p(B|A) \text{ from } p(A|B)$)
- $p(B|A) \text{ and } p(A|B)$ are conditional probabilities
- $p(A) \text{ and } p(B)$ are marginal probabilities [@lesaffre]


## Frequentist vs. Bayesian Approach

- The **Bayesian** Approach
    - Bayesian Inference can be written simply [@Koehrsen2018]
    
$$
Posterior = \frac{Likelihood \times Prior}{Normalization}
$$

- The $Prior$ is model of prior knowledge on the subject
- The $Likelihood$ is the probability of the data given the prior
- The $Normalization$ is a constant that ensures the posterior distribution is a valid density function whose integration is equal to 1
- The $Posterior$ is the probability model that expresses an updated view of the model parameters
  - From the initial parameters of the prior
  - Updated with new data expressed in the likelihood function


## Frequentist vs. Bayesian Approach

- The **Bayesian** Approach
  - A more formal expression of Bayes Rule applied for continous parameters
  
$$
\begin{align*}
p(\theta|y) =& \frac{ L(\theta|y)p(\theta) }{p(y)}\\
\\
p(\theta|y) \propto &   \text{ }L(\theta|y)p(\theta)
\end{align*}
$$

- The normalization constant ($p(y)$ above) ensures the posterior distribution is a valid distribution
  - The posterior density function can be written without this constant
- The resulting prediction is not a point estimate, but a distribution [@Bayes1991]

## The Bayesian Approach

- The Bayesian Linear Regression Model
  - changes based on:
      - distribution chosen for regression
      - distribution and hyperparameters chosen for priors
  - Our test case models a continuous outcome by a continuous predictor, so we use a Normal model with conjugate normal priors


## Role of prior knowledge in shaping predictions

- Priors can be subjective or objective
    - objective is preferred
- Noninformative priors can be used when there is not adequate prior
    knowledge
- Discounted priors are the result of adjusting a known prior to
    better reflect the current data[@lesaffre]

![Figure from [@lesaffre]](images/priorposterior.png)

## Understanding the Bayesian Framework

::: {layout-ncol=2}

- Bayes’ theorem is used to update prior beliefs about model
    parameters with new data
    - This results in a posterior distribution [@Bayes1991]
- Posterior distribution vs. point estimates
    - measures the uncertainty in predictions
    - richer picture for predictions
    - better uncertainty quantification [@Koehrsen2018]


![Figure from [@lesaffre]](images/priorposterior.png){fig-align="right" width="50%"}

:::

## Interpreting the Posterior

- The marginal posterior density function (output) may not be available
- Makov Chain Monte Carlo is commonly used
  - Markov chain sequence establishes a sample space from the posterior
  - Integration of samples generated through Monte Carlo techniques from the sample space
- Some popular MCMC algorithms:
  - Gibbs sampler
  - Metropolis-Hastings (MH) [@lesaffre]

## The Model

$$
\begin{align*}
Y_i|\beta_0, \beta_1, \sigma &\overset{\text{ind}}{\sim} N (\mu_i, \sigma^2) && \text{with } && \mu_i = \beta_0 + \beta_1X_i
\end{align*}
$$
Where:
- $Y_i$ is the arrival delay for the i-th flight
- $X_i$ is the departure delay for the i-th flight
- $\mu_i = \beta_0 + \beta_1X_i$ is the local mean arrival delay, ,  specific to the departure time 
- $\sigma^2$ is the variance of the errors
- $\overset{\text{ind}}{\sim}$ indicates conditional independence of each arrival delay with the given parameters

## Prior Selection

- Regression parameters
  - Intercept: $\beta_0 \sim N(m_0, s^2_0)$ 
  - Slope: $\beta_1 \sim N(m_1, s^2_1)$ 
  - Error: $\sigma \sim \text{Exp}(l)$

## The Bayesian Linear Regression Model

  The model can be written as

$$  
\begin{align*}
Y_i|\beta_0, \beta_1, \sigma &\overset{\text{ind}}{\sim} N (\mu_i, \sigma^2) && \text{with } && \mu_i = \beta_0 + \beta_1X_i \\ 
\beta_{0} &\sim N(m_0, s_0^2)\\
\beta_1 &\sim N(m_1, s_1^2)\\
\sigma &\sim \text{Exp}(l)
\end{align*}
$$

## Tuning Hyperparameters

$$
\begin{align*}
\beta_{0c} &\sim N(2, 36^2)\\
\beta_{1} &\sim N(0.02, 0.01^2)\\
\sigma &\sim \text{Exp}(0.02)
\end{align*}
$$

## The Updated Model

$$  
\begin{align*}
Y_i|\beta_0, \beta_1, \sigma &\overset{\text{ind}}{\sim} N (\mu_i, \sigma^2) && \text{with } && \mu_i = \beta_0 + \beta_1X_i \\ 
\beta_{0} &\sim N(2, 36^2)\\
\beta_1 &\sim N(0.02, 0.01^2)\\
\sigma &\sim \text{Exp}(0.02)
\end{align*}
$$

## Statistical Programming

- Data was analyzed in R [@R] , imported via CSV.
  - Data sourced from the [Bureau of Transportation Statistics](https://www.kaggle.com/datasets/patrickzel/flight-delay-and-cancellation-dataset-2019-2023?resource=download) via Kaggle @dataset.
- Libraries
  - rstanarm @rstanarm
    - stan_glm() function - simulation of model
    - posterior_predict() function - simulation of posterior
  - bayesrules @bayesrules
    - prediction_summary() function - evaluation of posterior
  
## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References

::: {#refs}
:::